% IEEE conference template — HPML Spring 2026 Final Report
% Team KV: Mamidibathula · Morisetty · Wang · Tian
% Columbia University
\documentclass[conference]{IEEEtran}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{microtype}

\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

% ----------------------------------------------------------------
\begin{document}

\title{Comparative Evaluation of KV Cache Optimization Strategies\\
for Efficient LLM Inference: Quantization, Sparse Attention,\\
Eviction, and Latent Projection}

\author{%
\IEEEauthorblockN{%
  Rishika Mamidibathula\textsuperscript{\dag},
  Suhas Morisetty\textsuperscript{*},
  Ziheng Wang\textsuperscript{*},
  Tony Tian\textsuperscript{*}}
\IEEEauthorblockA{%
  \textsuperscript{*}Department of Computer Science,
  \textsuperscript{\dag}Department of Data Science,
  Columbia University\\
  COMS E6998: High Performance Machine Learning, Spring 2026\\
  \{rm4318, vm2825, zw3153, jt3640\}@columbia.edu}
}

\maketitle

% ----------------------------------------------------------------
\begin{abstract}
Autoregressive large-language-model decoding is bounded by key-value
(KV) cache memory bandwidth: at batch size 32 and 4K context,
Llama-2-7B requires roughly 65\,GB for the cache alone, nearly
saturating an H100 80\,GB GPU.  We benchmark five KV optimization
techniques drawn from four compression families---per-token
quantization (KIVI 2/4-bit), dynamic sparse selection (TopK /
TokenSelect), one-shot eviction (SnapKV), and latent projection
(TransMLA)---together with a quantized-sparse hybrid (KIVI$\times$TopK),
under a single harness that fixes model, hardware, decode loop, prompts,
and seeds across runs.  KIVI 4-bit preserves baseline LongBench quality
and roughly doubles decode throughput at batch 32 by lifting the
operational batch ceiling from 32 to 128.  SnapKV slightly improves
overall quality with no per-step decode cost; TopK $K\!=\!1024$ matches
baseline quality and gains substantially on retrieval-heavy tasks.  MLA
yields a consistent $3.9\times$ KV reduction even at batch 1, but the
available checkpoint regresses on quality and requires retraining.  The
hybrid stacks memory savings of both compression axes at the cost of
decode speed at long context.  Each method occupies a distinct point in
the quality--memory--latency space; the right choice depends on whether
batch size, context length, or output fidelity is the binding
constraint.
\end{abstract}

\begin{IEEEkeywords}
KV cache, LLM inference, quantization, sparse attention, token eviction,
multi-head latent attention, throughput, memory efficiency
\end{IEEEkeywords}

% ================================================================
\section{Introduction}
% ================================================================

LLM inference proceeds in two phases.  Prefill processes the prompt in
a single forward pass; decode then emits one token per step.  Each
decode step rereads every cached key and value tensor from HBM, making
decode strongly memory-bandwidth bound rather than compute-bound.  The
cost grows linearly with sequence length, batch size, and depth.  For
Llama-2-7B \cite{llama2} (32 layers, 32 heads of dimension 128, FP16) at
4096 tokens and batch size 32, the cache occupies
$2\times32\times32\times128\times4096\times32\times2 \approx 65$\,GB
\cite{survey}---leaving little room for activations on an 80\,GB GPU
and pushing larger configurations into out-of-memory territory.

Research on KV reduction has converged on four principal compression
families.  \emph{Quantization} reduces bits per element while keeping
all tokens \cite{kivi,kvquant}.  \emph{Sparse selection} attends to a
dynamically chosen subset of tokens at each decode step
\cite{tokenselect,rocketkv}.  \emph{Eviction} permanently discards
low-importance tokens after prefill \cite{h2o,snapkv,adakv}.
\emph{Latent projection} stores a compressed low-rank representation
instead of full K/V tensors \cite{deepseekv2,transmla}.  Cross-family
comparison is difficult in practice because published evaluations
rarely share the same model, hardware, or decode loop.  This work
addresses that gap with a unified harness whose only variable is the KV
policy.

Our contributions are: (i) a single-codebase benchmark of five KV
optimization techniques and a quantized-sparse hybrid on Llama-2-7B;
(ii) a kernel-level analysis of crossover behavior---KIVI and TopK are
slower than baseline at low batch but faster at high batch or long
context; (iii) empirical evidence that KIVI 4-bit and SnapKV 0.4
deliver lossless or marginally better LongBench quality; and (iv) full
reproducibility via Modal cloud scripts and a public W\&B dashboard at
\url{https://wandb.ai/rm4318-columbia-university/kv-cache-hpml}.

% ================================================================
\section{Literature Review}
% ================================================================

\textbf{Quantization.}
KIVI \cite{kivi} stores keys with per-channel scales (capturing
channel-direction outliers) and values with per-token grouped
quantization (group size 32), keeping the last 128 tokens in FP16 for
short-range fidelity.  At 2-bit it achieves roughly $8\times$ cache
compression, at 4-bit roughly $4\times$.  KVQuant \cite{kvquant}
extends this with non-uniform codebooks and per-channel smoothing,
reaching near-lossless 1-bit at greater overhead.

\textbf{Sparse selection.}
TokenSelect \cite{tokenselect}---the basis of our TopK
implementation---retains the full FP16 cache but at each decode step
selects only the top-$K$ tokens for attention, fusing the scoring
kernel in Triton.  A cosine-similarity reuse cache avoids re-scoring
when consecutive queries are similar.  RocketKV \cite{rocketkv} adds a
two-stage compression pipeline; RefreshKV \cite{refreshkv} periodically
refreshes evicted tokens.

\textbf{Eviction.}
H2O \cite{h2o} identifies ``heavy-hitter'' tokens by cumulative
attention mass.  SnapKV \cite{snapkv} performs a single eviction after
prefill by voting over a small observation window, producing a
fixed-size compressed cache with no per-step overhead.  Ada-KV
\cite{adakv} allocates the budget adaptively per head.

\textbf{Latent KV.}
DeepSeek-V2 \cite{deepseekv2} introduced multi-head latent attention
(MLA), which stores a compressed latent vector $c_t$ per token and
expands to full K/V during attention via a learned up-projection, with
Attention Absorption amortizing the projection cost.  TransMLA
\cite{transmla} retrofits this structure onto an existing Llama-2
checkpoint.

% ================================================================
\section{Methodology}
% ================================================================

\subsection{Model and Hardware}

All experiments use \texttt{meta-llama/Llama-2-7b-chat-hf} (32 layers,
32 heads, head dim 128, FP16 weights) on a single NVIDIA H100 80\,GB
HBM3 GPU provided by Modal cloud compute.  The runtime stack is pinned:
PyTorch 2.4.1, HuggingFace Transformers 4.44.2, CUDA 12.1, Triton 3.0,
flash-attn 2.5.  The MLA experiments additionally use
\texttt{botsxc/llama2-7b-chat-mla-2048-8} \cite{transmla} (2048-dim
latent, 256-token calibration context, no further finetuning).

\subsection{Datasets and Metrics}

\textbf{Quality.}  LongBench \cite{longbench}, six tasks $\times$ 20
examples = 120 examples per method: \texttt{qasper} (scientific QA, F1),
\texttt{multifieldqa\_en} (multi-document QA, F1), \texttt{triviaqa}
(open-domain QA, F1), \texttt{2wikimqa} (multi-hop QA, F1),
\texttt{multi\_news} (summarization, ROUGE-L), \texttt{lcc} (code
completion, edit similarity).  Inputs are middle-truncated to 4096
tokens following the KIVI paper convention.

\textbf{Throughput.}  Synthetic fixed-size batches with prefill 1024
and generation 512 tokens, swept over batch sizes 1--256.  Five batches
per configuration; the first is dropped as warmup; the mean of the
remaining four is reported.

\textbf{Long context.}  Prefill 32{,}768 tokens, generation 512, batch
size 1, used for TopK-Flash and the KIVI$\times$TopK hybrid.

\subsection{Profiling Tools}

Memory: \texttt{torch.cuda.max\_memory\_allocated()} reset before each
batch; this measures peak, not steady-state.  Timing:
\texttt{torch.cuda.synchronize()} + \texttt{time.perf\_counter()}
brackets prefill (TTFT, time to first token) and the decode loop
(throughput in tokens per second over steps $2$ to $G$).
Operator-level traces are captured with \texttt{torch.profiler} and
exported to Chrome trace.  All runs are logged to Weights \& Biases.

\subsection{Method Implementations}

All methods implement a shared \texttt{MethodWrapper} interface in
\texttt{methods/} and run through an identical generate loop in
\texttt{benchmark/runner.py}.

\textbf{Baseline (FP16):} dense attention with no KV modification.

\textbf{KIVI \cite{kivi}.}  Keys are quantized per-channel to 2 or 4
bits (grouped across the sequence dimension per head); values are
quantized per-token with group size 32.  The last 128 tokens are kept
in FP16 (the residual buffer).  Quantization fuses INT4 packing in a
Triton kernel (\texttt{triton\_quantize\_and\_pack}); GEMV dequantizes
on-the-fly via \texttt{cuda\_bmm\_fA\_qB\_outer}.  The cache is stored
as a 9-tuple per layer (quantized blocks + scales + zeros for K and V
plus the FP16 residual).

\textbf{TopK / TokenSelect \cite{tokenselect}.}  The full FP16 cache is
retained.  Three-region attention budget: a sink window (first 128
tokens), dynamically-selected top-$K$ middle tokens, and a local window
(last 512 tokens).  Per-decode-step scoring fuses
$\mathbf{Q}\mathbf{K}^\top$ across all 32 heads in a Triton kernel
(\texttt{\_topk\_score\_kernel}) followed by softmax, per-head sum, and
\texttt{torch.topk}.  A cosine-similarity cache (threshold 0.9) skips
re-scoring when consecutive queries are similar; RoPE correction passes
\texttt{true\_seq\_length} to each step.

\textbf{TopK-Flash.}  The selection logic above on a paged KV pool
(256-token pages) plus \texttt{flash\_attn\_with\_kvcache} for
zero-copy selective attention.  This avoids the 52\,GB contiguous
allocation that standard flash-attn requires at 32K context.

\textbf{SnapKV \cite{snapkv}.}  After prefill, attention scores from
the last 32 query positions per head (the observation window) are
averaged, max-pooled with kernel size 7 to smooth noise, and the top
$\lfloor S \times 0.4 \rfloor$ tokens per head are retained
permanently.  Decode runs over the fixed reduced cache with no per-step
overhead.

\textbf{MLA (TransMLA \cite{transmla}).}  Each token stores a
compressed latent $c_t \in \mathbb{R}^{2048}$ ($4\times$ compression vs.\
$d_k\!\cdot\!n_h = 4096$).  At decode time, $c_t$ is up-projected to
full K/V by learned matrices; Attention Absorption folds the
up-projection into the QKV computation.

\textbf{KIVI$\times$TopK Hybrid.}  Prefill uses
\texttt{F.scaled\_dot\_product\_attention} (flash-attn backed,
$O(T)$ memory) to avoid 32K OOM.  Decode scores all tokens via
\texttt{cuda\_bmm\_fA\_qB\_outer} over INT4-packed K, selects
$n_\text{sink}+K+n_\text{local}$ positions, and gathers the
corresponding INT4 V rows into a contiguous buffer fed directly into
\texttt{cuda\_bmm\_fA\_qB\_outer}---no intermediate FP16 V is
materialized.

% ================================================================
\section{Experimental Results}
% ================================================================

\subsection{LongBench Quality}

Table~\ref{tab:quality} reports per-task and overall LongBench scores.
KIVI 4-bit and TopK $K\!=\!1024$ both achieve 0.292 (+0.3\% and +0.4\%
over baseline 0.291)---within evaluation noise for 20 examples per
task but consistent across runs.  SnapKV 0.4 reaches 0.295 (+0.3\%).

\begin{table}[!t]
\centering
\caption{LongBench quality (F1 / ROUGE-L / edit-sim), 6 tasks $\times$ 20 ex.
Tasks: (1) qasper, (2) mfqa, (3) triviaqa, (4) 2wikimqa,
(5) mnews, (6) lcc.}
\label{tab:quality}
\setlength{\tabcolsep}{3.5pt}
\begin{tabular}{@{}lrrrrrrr@{}}
\toprule
\textbf{Method} & \textbf{(1)} & \textbf{(2)} & \textbf{(3)} & \textbf{(4)} & \textbf{(5)} & \textbf{(6)} & \textbf{Avg} \\
\midrule
Baseline (FP16)       & .172 & .380 & .644 & .083 & .181 & .287 & .291 \\
KIVI 4-bit            & .194 & .360 & .643 & .083 & .188 & .286 & \textbf{.292} \\
KIVI 2-bit            & .133 & .382 & .608 & .050 & .191 & .277 & .273 \\
TopK $K$=2048         & .145 & .329 & .731 & .075 & .178 & .282 & .290 \\
TopK $K$=1024         & .153 & .281 & .773 & .089 & .183 & .273 & \textbf{.292} \\
TopK $K$=512          & .145 & .237 & .751 & .050 & .178 & .281 & .274 \\
SnapKV 0.4            & .153 & .387 & .687 & .083 & .181 & .277 & \textbf{.295} \\
MLA (B=1)*            & .036 & .101 & .064 & .043 & .103 & .140 & .081 \\
\bottomrule
\end{tabular}
\vspace{1pt}
\footnotesize{* MLA baseline truncated to 2048 tokens vs.\ 4096 for others.}
\end{table}

Per-task patterns reveal two opposite breakdown mechanisms when
compression is too aggressive.  KIVI 2-bit and TopK $K\!=\!512$ both
end up near 0.274 overall, but for different reasons.  KIVI's
quantization noise hurts qasper ($-22\%$) and 2wikimqa ($-40\%$), tasks
that require precise recall of sparse mid-context facts.  TopK
$K\!=\!512$ hurts \texttt{multifieldqa\_en} ($-38\%$) because answer
spans deeper than $n_\text{sink} + n_\text{local}$ are pruned.
Conversely, TopK beats baseline on TriviaQA across all $K$
($+13$\,pt to $+20$\,pt): the answer appears in multiple paraphrases,
so sparse selection drops distracting copies and concentrates
attention.  SnapKV's one-shot vote actively aligns selection with
answer-bearing positions, boosting triviaqa ($+4$\,pt) and multifieldqa
($+0.7$\,pt) while only hurting qasper ($-2$\,pt) where evidence spans
fall outside the 32-token observation window.

\subsection{Throughput and Memory}

Table~\ref{tab:throughput} shows the KIVI throughput sweep at prefill
1024, generation 512.

\begin{table}[!t]
\centering
\caption{Decode throughput (tok/s) and peak GPU memory vs.\ batch size.
Prefill 1024, gen 512.  ``--'' = OOM.}
\label{tab:throughput}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{@{}rrrrrrr@{}}
\toprule
 & \multicolumn{3}{c}{\textbf{Decode tok/s}} & \multicolumn{3}{c}{\textbf{Peak mem (GB)}} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7}
\textbf{BS} & \textbf{Base} & \textbf{KIVI-2b} & \textbf{Spdup} & \textbf{Base} & \textbf{KIVI-2b} & \textbf{MLA} \\
\midrule
1   &  70 &   30 & 0.43$\times$ & 15.1 & 14.0 & 13.8 \\
2   & 135 &   61 & 0.45$\times$ & 16.7 & 14.4 &  --   \\
4   & 242 &  123 & 0.51$\times$ & 20.0 & 15.3 &  --   \\
8   & 351 &  243 & 0.69$\times$ & 26.4 & 17.1 &  --   \\
16  & 438 &  480 & 1.10$\times$ & 39.3 & 20.7 &  --   \\
32  & 497 &  957 & 1.93$\times$ & 65.0 & 27.8 &  --   \\
64  &  -- & 1295 & $\infty$     &  --  & 42.1 &  --   \\
128 &  -- & 1517 & $\infty$     &  --  & 70.7 &  --   \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Crossover at BS=16.}  KIVI 2-bit decode throughput is
$0.4$--$0.7\times$ baseline at BS 1--8 but crosses over at BS=16
(480 vs.\ 438\,tok/s, $1.10\times$) and reaches $1.93\times$ at BS=32.
The mechanism is straightforward: at small batch the decode
$\mathbf{Q}\mathbf{K}^\top$ is a GEMV, and INT4 GEMV does not map to
H100 tensor-core units---it is roughly $2.5\times$ slower than FP16
GEMV.  At large batch the same operation becomes a near-GEMM where the
HBM bandwidth saving (the cache is $2.3\times$ smaller at BS=32:
27.8 vs.\ 65.0\,GB) outweighs dequantization cost.

\textbf{Capacity extension.}  Baseline OOMs at BS=64; KIVI 2-bit
reaches BS=128 at 70.7\,GB and achieves 1{,}517\,tok/s---about
$3\times$ the maximum ever achievable with the baseline on this GPU.
This is the strongest practical argument for KIVI: it raises the
per-GPU concurrency ceiling, not just speed at fixed batch.

\textbf{TTFT overhead.}  KIVI's TTFT is 18--66\% worse across batch
sizes (998 vs.\ 1659\,ms at BS=32) because the full KV cache is
quantized and packed at the end of prefill---a cost proportional to
$B \times S$.  For latency-sensitive interactive workloads this is a
real penalty; for throughput-oriented offline inference it is amortized
over hundreds of decode steps.

\textbf{MLA at B=1.}  At gen=256, MLA achieves 43--44\,tok/s vs.\
34\,tok/s baseline ($1.26$--$1.32\times$) across prefill lengths
512--2048.  Even single-request decode is partially HBM-bandwidth
bound: the $3.9\times$ KV reduction (e.g., 1207 \(\rightarrow\) 311\,MB
at $p\!=\!2048$) reduces per-step HBM traffic proportionally.  The
speedup narrows at gen=512 because the MLA up-projection cost grows
with output length.

\subsection{Long-Context Results (32K)}

\begin{table}[!t]
\centering
\caption{Long-context results.  Prefill 32{,}768 tokens, gen 512, BS=1.}
\label{tab:longctx}
\begin{tabular}{@{}lrrrr@{}}
\toprule
\textbf{Method} & \textbf{TTFT (ms)} & \textbf{Spdup} & \textbf{Peak mem} & \textbf{Saving} \\
\midrule
Baseline (FP16)     & 2627 & 1.00$\times$ & 52.1\,GB & --- \\
TopK-Flash          & 1761 & 1.49$\times$ & 42.1\,GB & $-19\%$ \\
SnapKV 0.4          &  --- & ---          & 26.0\,GB$^{\dagger}$ & $-50\%$ \\
KIVI$\times$TopK 4b &  --- & 0.61$\times$ & 25.8\,GB & $-50\%$ \\
\bottomrule
\end{tabular}
\vspace{1pt}
\footnotesize{Baseline OOMs at 64K; TopK-Flash succeeds at 65.8\,GB.
$^{\dagger}$SnapKV peak memory at 32K is interpolated from
short-context measurements; 32K SnapKV throughput not directly
benchmarked.}
\end{table}

\textbf{TopK-Flash.}  At 32K context the 256-token paged KV pool
eliminates the 52\,GB contiguous allocation that standard flash-attn
requires, reducing TTFT by 33\% and peak memory by 19\%.  The win is
not from sparsity but from allocation and memory-access-pattern
improvement.  Baseline OOMs at 64K; TopK-Flash succeeds at 65.8\,GB.

\textbf{KIVI$\times$TopK hybrid.}  Combining INT4 storage with sparse
attention cuts peak memory 50\% (52.1$\rightarrow$25.8\,GB) but decode
throughput drops 39\% (21.0$\rightarrow$12.9\,tok/s).  The slowdown has
two sources: (a) $\mathbf{Q}\mathbf{K}^\top$ must still score all 32K
tokens every step before top-$K$ can be applied---the scoring cost is
identical to baseline, plus topk + sort + gather; (b) the INT4 V GEMV
inherits the same low-batch dequantization penalty as KIVI.  LongBench
quality remains at 0.297, within noise of 0.291.  The hybrid shows
that orthogonal compression strategies stack memory savings, but a
selection-cost reduction is needed before the latency win materializes.

% ================================================================
\section{Discussion}
% ================================================================

\subsection{Deployment Recommendations}

Table~\ref{tab:tradeoff} summarizes the operating point of each method.

\begin{table}[!t]
\centering
\caption{Deployment fit summary.}
\label{tab:tradeoff}
\setlength{\tabcolsep}{3pt}
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Method} & \textbf{KV ratio} & \textbf{Quality} & \textbf{Best regime} \\
\midrule
KIVI 4-bit  & $4\times$  & Lossless (.292)        & Batch serving $\geq$ BS=16 \\
KIVI 2-bit  & $8\times$  & $-6.2\%$ (.273)        & Max-batch, accept noise \\
TopK $K$=1024 & $1\times$$^{\dagger}$ & Lossless (.292) & Single-req, long context \\
SnapKV 0.4  & $2.5\times$ & $+0.3\%$ (.295)       & Any; quality-neutral \\
MLA (B=1)   & $3.9\times$ & $-65\%$* needs retrain & After finetuning at $\geq$4K \\
\bottomrule
\end{tabular}
\vspace{1pt}
\footnotesize{$^{\dagger}$TopK keeps full FP16 cache; only attention scope is sparse.  *Truncation mismatch may inflate gap.}
\end{table}

\textbf{KIVI 4-bit} is the best practical default for high-throughput
batch inference: it costs nothing in quality, raises max batch from 32
to 128, and roughly triples peak single-GPU throughput.  The cost is
TTFT latency ($+18$--$66\%$), which matters for interactive
applications.

\textbf{SnapKV 0.4} is the right choice when quality must be preserved
and per-step overhead must be zero.  Its $2.5\times$ KV reduction has
no decode-time cost because the compressed cache is built once after
prefill.

\textbf{TopK $K\!=\!1024$} is suited to single-request, long-context
use ($>32$K) where the paged variant (TopK-Flash) provides real TTFT
and memory gains.  At short context and low batch, scoring overhead
dominates the saved attention work.

\textbf{MLA} shows the theoretical ceiling of latent projection:
$3.9\times$ KV reduction and $1.3\times$ decode speedup are repeatable
benefits even at B=1, but production-quality output requires a
checkpoint trained at longer context with higher latent rank.

\subsection{Kernel-Level Observations}

\textbf{INT4 GEMV inefficiency.}  \texttt{cuda\_bmm\_fA\_qB\_outer}
performs on-the-fly INT4 $\rightarrow$ FP16 dequantization per GEMV.
H100 tensor cores target INT8/FP16 GEMM; INT4 GEMV falls back to scalar
or vector-INT4 paths.  Bandwidth saving only exceeds this overhead when
decode approaches GEMM at BS$\geq$16.

\textbf{KIVI residual buffer hot path.}  At short generation lengths
($<\!\texttt{residual\_length}\!\times\!n_\text{layers}$), nearly all
attention concentrates in the FP16 residual; KIVI provides minimal
storage benefit in this regime.  This is also why KIVI TTFT is roughly
$3\times$ slower: the full prefill quantization is paid upfront.

\textbf{Triton fused scoring.}  \texttt{\_topk\_score\_kernel} fuses
$\mathbf{Q}\mathbf{K}^\top$ + softmax + head-sum into a single launch,
reducing Python-dispatch overhead $2$--$3\times$ vs.\ sequential
PyTorch ops.  At 1K--4K sequence lengths the critical path remains HBM
bandwidth, not compute or launch latency.

\textbf{Hybrid stays in INT4.}  The hybrid keeps data in INT4 across
the gather $\rightarrow$ attention boundary by passing the gathered
packed tensor directly into \texttt{cuda\_bmm\_fA\_qB\_outer},
avoiding an FP16 V materialization that would have cost an extra
allocation and HBM round-trip per layer per step.

\subsection{Limitations}

\emph{Single architecture.}  Results are specific to Llama-2-7B with
multi-head attention (32 heads, 32 KV heads).  Models with grouped-query
attention (Llama-3, Mistral) have fewer KV heads, which changes the
compression profile.

\emph{Small evaluation set.}  LongBench results use 20 examples per
task.  Some per-task deltas (e.g., qasper KIVI 2-bit $-22\%$) may
partially reflect small-$n$ variance, although the direction is
consistent across tasks.

\emph{MLA checkpoint.}  The available TransMLA checkpoint was
calibrated at 256-token context without task finetuning; a properly
trained checkpoint would likely recover much of the 65\% quality gap.

\emph{Hybrid scoring cost.}  The 50\% memory win comes with a 39\%
decode penalty because scoring must precede selection.  Hierarchical
or lazy-scoring approaches could decouple them.

% ================================================================
\section{Conclusion}
% ================================================================

We conducted a controlled benchmark of five KV-cache optimization
methods spanning four compression families on Llama-2-7B.  KIVI 4-bit
provides lossless quality with $4\times$ KV compression and
$1.93\times$ throughput at BS=32---our recommended default for
batch-oriented serving.  SnapKV marginally improves LongBench overall
score with zero runtime overhead.  TopK $K\!=\!1024$ matches baseline
quality and gains substantially on retrieval-heavy tasks.  MLA delivers
a consistent $3.9\times$ KV reduction and $1.3\times$ decode speedup at
B=1 but requires a properly trained checkpoint for production quality.
The KIVI$\times$TopK hybrid stacks memory savings of both axes but pays
a decode penalty until scoring cost can be reduced.  No single method
dominates all axes; practitioners should select based on whether batch
size, context length, or output fidelity is the binding constraint.

Future work should address: (i)~batched decode for TopK to amortize the
per-step scoring cost and produce a KIVI-style crossover at
BS$\geq$8; (ii)~MLA retraining at 4K--8K context with rank 16--32 to
preserve $3.9\times$ compression while recovering QA quality;
(iii)~hierarchical or lazy scoring in the KIVI$\times$TopK hybrid to
decouple the memory win from the decode penalty; and (iv)~cross-method
composition, e.g., SnapKV eviction followed by KIVI quantization on the
retained tokens to stack savings from orthogonal axes.

\appendix
\section*{AI Use Disclosure}

Per the HPML AI Use Policy.  Team members used Claude (Anthropic) and
GitHub Copilot for: debugging CUDA OOM errors during KIVI kernel
integration, clarifying Triton kernel semantics, and polishing report
prose drafted by team members.  All experimental numbers were produced
by running our own Modal scripts on H100 containers; profiler-trace
interpretations were verified against raw \texttt{torch.profiler}
traces in \texttt{results/}.  AI-suggested code was rewritten in our
own words and re-validated to produce identical outputs to a hand-written
reference.  No analysis, interpretation, or written argument in this
report was generated end-to-end by an AI tool.

% ================================================================
\bibliographystyle{IEEEtran}

\begin{thebibliography}{14}

\bibitem{survey}
J.~Jiang, P.~Yang, R.~Zhang, and F.~Liu,
``Towards efficient large language model serving: A survey on
system-aware KV cache optimization,''
\textit{TechRxiv}, 2026.

\bibitem{llama2}
H.~Touvron \textit{et al.},
``Llama 2: Open foundation and fine-tuned chat models,''
\textit{arXiv:2307.09288}, 2023.

\bibitem{kivi}
Z.~Liu \textit{et al.},
``KIVI: A tuning-free asymmetric 2-bit quantization for KV cache,''
in \textit{Proc.\ ICML}, 2024.

\bibitem{kvquant}
C.~Hooper \textit{et al.},
``KVQuant: Towards 10 million context length LLM inference with KV
cache quantization,''
in \textit{NeurIPS}, 2024.

\bibitem{tokenselect}
W.~Wu \textit{et al.},
``TokenSelect: Efficient long-context inference via dynamic token-level
KV cache selection,''
in \textit{Proc.\ EMNLP}, 2025.

\bibitem{rocketkv}
P.~Behnam \textit{et al.},
``RocketKV: Accelerating long-context LLM inference via two-stage KV
cache compression,''
in \textit{Proc.\ ICML}, 2025.

\bibitem{refreshkv}
F.~Xu, T.~Goyal, and E.~Choi,
``RefreshKV: Updating small KV cache during long-form generation,''
in \textit{Proc.\ ACL}, 2025.

\bibitem{h2o}
Z.~Zhang \textit{et al.},
``H2O: Heavy-hitter oracle for efficient generative inference of large
language models,''
in \textit{NeurIPS}, 2023.

\bibitem{snapkv}
Y.~Li \textit{et al.},
``SnapKV: LLM knows what you are looking for before generation,''
in \textit{NeurIPS}, 2024.

\bibitem{adakv}
Y.~Feng \textit{et al.},
``Ada-KV: Optimizing KV cache eviction by adaptive budget allocation,''
in \textit{NeurIPS}, 2025.

\bibitem{deepseekv2}
DeepSeek-AI,
``DeepSeek-V2: A strong, economical, and efficient mixture-of-experts
language model,''
\textit{arXiv:2405.04434}, 2024.

\bibitem{transmla}
Y.~He \textit{et al.},
``TransMLA: Multi-head latent attention is all you need,''
\textit{arXiv:2502.07864}, 2025.

\bibitem{xkv}
C.-C.~Chang \textit{et al.},
``xKV: Cross-layer SVD for KV-cache compression,''
\textit{arXiv:2503.18893}, 2025.

\bibitem{longbench}
Y.~Bai \textit{et al.},
``LongBench: A bilingual, multitask benchmark for long context
understanding,''
\textit{arXiv:2308.14508}, 2023.

\end{thebibliography}

\end{document}
